### 2015-07-20 background-new.tex
a finished version: jul 20, 11 AM

\chapter{Traffic engineering in content-dominated networks: background and motivation}
\label{ch:te-background}

This chapter serves two main purposes. First is to provide the reader with necessary background on ISP traffic engineering, content delivery and the interaction between traffic engineering and application-layer traffic adaptation (Section \ref{sec:bg-bg}). Second is to explain the motivation for our work on traffic engineering in content-dominated networks in light of the existing work in this area (Section \ref{sec:bg-motivation}).

@@

%This chapter provides background on content delivery and traffic engineering and makes a case for studying the role of content placement on this interaction. Chapter x and Chapter y studies traffic engineering in two scenarios with varying degrees of flexiblity in placing content. Chapter x focuses on an ISP network in a fixed content placement policy, namely that of placing content at multiple locations chosen randomly. While, Chapter y focuses on a network CDN, an ISP that deploys a CDN to deliver content to users on its network and enjoys full freedom to control the placement and routing on its network.

@@

\begin{itemize}
	\item ISP traffic engineering schemes compute routing for optimizing link-utilization based cost functions. These schemes commonly take a demand-aware approach that uses previously measured traffic matrics for computing future network routes (Section \ref{sec:ch2-te}).	
	\item CDNs commonly use \emph{demand-oblivious} techniques, i.e., techniques that require no system-wide measurement of content demand, for content placement and request redirection towards improving user-perceived performance across the Internet (Section \ref{sec:ch2-cdn}).
	\item Interaction between traffic engineering and application-layer adaptation affects user-perceived performance as well as network cost. But, prior work has focused mostly on the interaction of traffic engineering with two aspects of application-layer adaptation --  overlay routing and request redirection (Section \ref{sec:ch2-te-cdn}).
\end{itemize}

@@

A key goal of traffic engineering is to avoid congestion hotspots in the network by optimizing routes based on network topology and expected traffic demand that is represented in the form of traffic matrix. In ISP networks, traffic engineering decides both intra-domain routing (within the ISP) and inter-domain routing (across ISPs). We focus here on intra-domain routing and refer the reader to  \cite{Feamster2003,rexford} for a survey of inter-domain traffic engineering.

The evaluation metric for traffic engineering is a cost function that is dependent on the utilization of network links. Link utilization is a relevant metric because a low utilization of all links implies that the network is free from congestion hotspots and has spare capacity to tolerate an increase in demand. A well-known cost function is maximum link utilization or MLU.

We give a simple example to show how traffic engineering can reduce link utilization. In Figure \ref{fig:te-example}, there are two links of capacity 1 Mbps and 3 Mbps between nodes A and B. The traffic from A to B is 2 Mbps. If a shortest path routing is followed, all traffic must be sent through either of the two links. The least MLU = 0.67 is achieved by using the large capacity link. A more flexible routing approach is to split traffic among the two links. Such a flow-split routing achieves MLU = 0.5 by sending 0.5 Mbps and 1.5 Mbps via the two links respectively. Thus, a better engineered routing resulted in a lower MLU in this example.

@@

If the traffic matrix is known accurately, the optimal solution to the traffic engineering problem can formulated as a multi-commodity flow optimization. The routing thus computed is refered to as optimal traffic engineering in literature. As it is impossible to have accurate knowledge of future traffic matrices, traffic engineering schemes either take a \emph{demand-aware} or a \emph{demand-oblivious} approach. A demand-aware  traffic engineering  periodically updates routing using historically observed traffic matrices. A demand-oblivious  traffic engineering  requires no explicit measurement of traffic matices, but instead configures routing statically. A demand-oblivious  traffic engineering  is simpler to implement because it requires neither measuremnt nor periodic updating of network routing.

@@

Content delivery systems seek to provide a high-quality experience to users accessing content in all regions  at all times. A canonical example of a content delivery system is a content delivery network (CDN). State-of-the-art CDNs operate geo-distributed datacenters, and use a combination of edge caching, intelligent request redirection, and path and protocol optimizations for delivery of several types of content, e.g., video, bulk downloads, and interactive websites \cite{DilleyMPPSW02,akamai-overview}. Given their geo-distributed deployment, the decisions of content placement, i.e., locations at which a content is placed, and request redirection, i.e., which location is best positioned to serve a user's request, are central to the functioning of a CDN.

\textbf{Content placement:} Content placement in CDNs is commonly done using caching schemes. For example a commonly used caching strategy is least recently used (LRU) cache replacement. There are two reason why content cahing is widely used. First, caching naturally captures geographic and temporal locality in content requests to populate caches with content likely to be reused. Second, a vast majority of network traffic is generated by content that gets updated infrequently, e.g., a video, audio, images. As a result, cached copies of content remain reusable for long duration.

@@

Content delivery techniques can be classified into demand-aware and demand-oblivious similar to our classfication of traffic engineering schemes. We consider common content delivery techniques discussed above to be demand-oblivious since they do not require system-wide measurement of content demand. In contrast, a demand-aware approach to  placement and redirection has also been studied, e.g., Applegate \emph{et al.} use a demand-aware approach to determine placement and redirection for Video-on-Demand content in the network.

@@

\textbf{Interaction between traffic engineering and overlay routing:} Several results show the negative interaction between selfish overlay routing and network routing \cite{Roughgarden,selfishQiu}. Theoretical results indicate that the negative interaction could cause an arbitrary degradation in user perceived delay. Further studies using ``Internet-like'' traffic demands and topologies indicate that this interaction hurts traffic engineering metrics. In this thesis, we do not consider overlay routing since our focus is primarily on intra-domain routing while overlay routing is most commonly used to mask inter-domain routing inefficiencies.

%However, it appears that a small fraction of CDN traffic uses overlay routing.  For example, traffic from CDN edge server to the client always follows network routing. Further, overlay routing yields ``marginal" benefits ($<$ 20\%) over network routing for 79\%-96\% of paths depending on which geographic region is being considered  \cite{rahul2006overlays}, which suggests that traffic between CDN servers forming an overlay network follows network routing in most cases. For this reason, this thesis does not model the interaction between overlay and network routing \tbd{is this the best reason?}.

\textbf{Interaction between traffic engineering and request redirection:} Recent research has investigated the interaction between request redirection and traffic engineering, without considering the role of content placement. This interaction is commonly studied in the context of Internet service providers (ISPs) and content providers (CPs) with geo-distributed datacenters. Both analytical results \cite{Jiang2009,JohariGameTheory} and system implementations \cite{CATE,P4P} have shown that there is value for joint optimization of request redirection and traffic engineering, and cooperative strategies can help traffic engineering metrics while maintaining or improving user-perceived performance. Our work distinguishes in that  we model the three-way interaction between placement, redirection and routing and show that placement flexibility is more powerful degree of freedom than redirection or routing to improve traffic engineering metrics as well as to reduce user-perceived latency.

@@

A content-dominated traffic and the presence of content at multiple network locations changes the traditional ISP traffic enginering problem. In such a network, a traffic matrix is no longer an effective representation of demand. It can be changed via application-layer adaptation without any underlying change in demand (Section \ref{sec:bg-3node}). Demand is better expressed as a \emph{content matrix} in which each entry represents the demand for a content at a particular network location. A traffic matrix implicitly assumes a fixed source and a fixed destination for each traffic flow. But, a content matrix expresses the fact that the demand for a content at a particular destination can be served from any set of source locations. As we explain using examples below, the flexibility of serving the demand from any location raises questions on the usefulness of link-utilization based metrics that are commonly used in traffic engineering (Section \ref{sec:bg-poormlu}) as well on the importance of traffic engineering itself (Example \ref{sec:bg-ncdn-interaction}). These and other related questions motivate our work on traffic-engineering in content dominated networks.

@@

\subsection{ISP network with content location diversity}
\label{sec:bg-locdiv}
We consider an ISP network that has little control over placement and redirection, but applications can leverage \emph{location diversity}, or  the ability to download content from multiple locatioons. We motivate why traffic engineering schemes must be evaluated while accounting for the effect of location diversity in the network, and why we need to look beyond link-utilization based metrics in evaluating traffic engineering schemes. These questions are addressed in Chapter \ref{ch:beyondmlu}.

@@

We explain why link-utilization based metrics may not reflect user-perceived performance. Further, they are a poor metric to measure network capacity, i.e., the factor of surges in traffic demand that a network can tolerate, in a network with location diversity.

Link-utilization based metrics may not be a good predictor of user-perceived performance that depends on other factors such as propagtion delay, access link capacity and backbone link capacity also. Moreover, queuing-theoretic models show that only a high link utilization severely affects performance-critical metrics such as queuing delay and packet loss. For low to moderate link utilization that is common in today's ISPs, a reduction in link utilization may not yield a commensurate benefit in application peformance. This observation has also been confirmed by explicit measurements on Internet backbones \cite{ExpRouterBuffer}, and is consistent with studies on ISP backbones showing that over 90\% of all packet loss is caused by interdomain routing fluctuations as opposed to high utilization \cite{SprintStudy} and 90\% of TCP flows experience no packet loss \cite{SprintBackbone}. Furthermore, end-to-end Internet path delays are known to be largely determined by propagation delays as opposed to queueing delays \cite{SprintBackbone,SingleHopDelay}.

@@

Link-utilization based metrics such as MLU are poor metric for network capacity, i.e., the factor of surges in traffic demand that a network can tolerate, in a network with location diversity. MLU provides a simple intuitive capacity metric, the inverse of the MLU serves as a metric of capacity, e.g., if a traffic engineering scheme achieves an MLU of 0.25 for a given matrix, then it can tolerate up to a 4$\times$ surge in the load represented by the matrix. Unfortunately, as the example in Figure \ref{fig:3node-bg} shows, MLU is not a meaningful metric of capacity when application adaptation to location diversity determines the traffic matrix. Thus, location diversity necessitates a new capacity metric.

@@

\subsection{Network CDN}
\label{sec:bg-ncdn}
Network CDNs are a recent, potentialy transformative trend, in which ISPs deploy CDNs on their infrastructures.  Unlike traditional ISPs and traditional CDNs, NCDNs enjoy full control over placement, redirection and routing on their networks. The interaction between placement and routing in an NCDN illustrated below motivates us to explore the benefit of joint optimization of placement, redirection and routing as well as to evaluate the relative importance of placement and routing in an NCDN. These questions are addressed in Chapter \ref{ch:ncdn}.

@@

To appreciate how placement can shape traffic in an NCDN, consider the simple example in Figure~\ref{fig:NetworkExample}. Node $C$ has an object in its cache that is requested by end-users at nodes $A$ and $D$. Suppose that one unit of traffic needs to be routed from $C$ to $A$ and $0.5$ units  from $C$ to $D$ to satisfy the demand for that object. The routing that achieves the minimum MLU of $0.5$ to serve the demanded object is shown in the figure. Note that the routing that achieves the MLU of 0.5 is not possible with a simple, \unplanned\ protocol like InverseCap as that would route all the traffic demand from $C$ to $A$ via $B$, resulting in an MLU of 1. Thus, a (\planned) traffic engineering scheme is necessary to achieve an MLU of 0.5.

@@

On the other hand, NCDNs can shape the traffic demand matrix by using a judicious placement and redirection strategy. Suppose that there is some space left in the content server's cache at node $B$ to accommodate an additional copy of the demanded object. By creating an additional copy of the object at $B$, the traffic demand of $A$ can be satisfied from $B$ and the demand of $D$ from $C$ achieving an MLU of $0.125$. In this case, judicious content placement decreased the MLU by a factor of $4$. Even more interestingly, this best MLU can be achieved using a simple routing scheme like hop-count routing while also improving user-perceived latency (assuming that the latency of link $BA$ is lower than that of the two-hop paths from $C$ to $A$). While this is a toy example, it shows that content placement flexibility reduces network cost and enables simpler routing. This interaction between placement and routing motivates us address the following questions for real content workloads and network topologies.

>>>>

\chapter{Traffic engineering in content-dominated ISP networks: background and motivation}
\label{ch:te-background}

This chapter serves two main purposes. First is to provide the reader with necessary background on ISP traffic engineering, content delivery and the interaction between traffic engineering and application-layer traffic adaptation (Section \ref{sec:bg-bg}). Second is to explain the motivation for our work on traffic engineering in content-dominated ISP networks in light of the existing work in this area (Section \ref{sec:bg-motivation}).

@@

%This chapter provides background on content delivery and traffic engineering and makes a case for studying the role of content placement on this interaction. Chapter x and Chapter y studies traffic engineering in two scenarios with varying degrees of flexiblity in placing content. Chapter x focuses on an ISP network in a fixed content placement scheme, namely that of placing content at multiple locations chosen randomly. While, Chapter y focuses on a network CDN, an ISP that deploys a CDN to deliver content to users on its network and enjoys full freedom to control the placement and routing on its network.

@@

\begin{itemize}
	\item 
	ISP traffic engineering schemes compute routing for optimizing link-utilization based cost functions. These schemes commonly take a demand-aware approach that uses previously measured traffic matrics for computing future network routes (Section \ref{sec:ch2-te}).	
	\item 
	CDNs commonly use \emph{demand-oblivious} content placement and request redirection techniques towards improving user-perceived performance across the Internet (Section \ref{sec:ch2-cdn}). These techiques requires do not require content-level measurement of demand but instead use simple online heuristics to make their decision.
	\item 
	Prior work on the interaction between traffic engineering and application-layer adaptation focuses primarily on two aspects of application-layer adaptation --  overlay routing and request redirection. These interactions result in globally sub-optimal network cost as well as user-perceived latency, and co-operative mechanisms can leverage these interactions to improve cost and performance metrics.
\end{itemize}

@@

A key goal of ISP traffic engineering is to avoid congestion hotspots in the network by optimizing routes based on network topology and expected traffic demand that is represented in the form of traffic matrix. In ISP networks, traffic engineering decides both intra-domain routing (within the ISP) and inter-domain routing (across ISPs). We focus here on intra-domain routing and refer the reader to  \cite{Feamster2003,rexford} for a survey of inter-domain traffic engineering.

The evaluation metric for ISP traffic engineering is a cost function that is dependent on the utilization of network links. A well-known cost function is maximum link utilization or MLU. A low link utilization is desirable to ISPs for two reasons. First, a low utilization of all links implies that the network is free from congestion hotspots. Second, it also implies that a network has more spare capacity to tolerate an increase in demand. Consider MLU as a capacity metric for example. If a traffic engineering scheme achieves an MLU of 0.25 for a given matrix, then it can tolerate up to a 4$\times$ surge in the load represented by the matrix.

We give an example to show how traffic engineering reduces link utilization. In Figure \ref{fig:te-example}, there are two links of capacity 1 Mbps and 3 Mbps between nodes A and B. The traffic from A to B is 2 Mbps. If a shortest path routing is followed, all traffic must be sent through either of the two links. The least MLU = 0.67 is achieved by using the large capacity link. A more flexible routing approach is to split traffic among the two links. Such a flow-split routing achieves MLU = 0.5 by sending 0.5 Mbps and 1.5 Mbps via the two links respectively. Thus, a better engineered routing resulted in a lower MLU in this example.

@@

If the traffic matrix is known accurately, the optimal solution to the traffic engineering problem can formulated as a multi-commodity flow optimization. The routing thus computed is refered to as \emph{optimal traffic engineering} in literature. As it is impossible to have accurate knowledge of future traffic matrices, traffic engineering schemes either take a \emph{demand-aware} or a \emph{demand-oblivious} approach. A demand-aware  traffic engineering  periodically updates routing using historically observed traffic matrices. A demand-oblivious  traffic engineering  requires no explicit measurement of traffic matices, but instead configures routing statically. A demand-oblivious  traffic engineering  is simpler to implement because it requires neither measuremnt nor periodic updating of network routing.

@@

Content delivery systems seek to improve user-perceived performance for content accesses in all regions  at all times. A canonical example of a content delivery system is a content delivery network (CDN). State-of-the-art CDNs operate geo-distributed datacenters, and use a combination of edge caching, intelligent request redirection, and path and protocol optimizations for delivery of several types of content, e.g., video, bulk downloads, and interactive websites \cite{DilleyMPPSW02,akamai-overview}. Given their geo-distributed deployment, the decisions of content placement, i.e., locations at which acontent is placed, and request redirection, i.e., which location is best positioned to serve a user's request, are central to the functioning of a CDN.

\textbf{Content placement:} Content placement in CDNs is commonly done using caching schemes. For example a commonly used caching scheme is least recently used (LRU) cache replacement. There are two reason why content cahing is widely used. First, caching naturally captures geographic and temporal locality in content requests to populate caches with content likely to be reused. Second, a vast majority of network traffic is generated by content that gets updated infrequently, e.g., a video, audio, images. As a result, cached copies of content remain reusable for long duration.

@@

Content delivery techniques can be classified into demand-aware and demand-oblivious similar to our classfication of traffic engineering schemes. We consider common content delivery techniques discussed above to be demand-oblivious since they do not require long term content-level measurement of demand but instead use simple online heuristics to make their decision.. In contrast, a demand-aware approach to  placement and redirection has also been studied, e.g., Applegate \emph{et al.} use a demand-aware approach to determine placement and redirection for Video-on-Demand content in the network.

@@

\textbf{Interaction between traffic engineering and overlay routing:} Several results show the negative interaction between selfish overlay routing and network routing \cite{Roughgarden,selfishQiu}. Theoretical results indicate that the negative interaction could cause an arbitrary degradation in user perceived delay. Further studies using Internet-like traffic demands and topologies indicate that this interaction hurts traffic engineering metrics. While this body of work focuses on application adaptation to \emph{path diversity} created by overlay routing, our work focuses on \emph{location diversity} present in the Internet (Chapter \ref{ch:beyondmlu}). We do not consider overlay routing since our focus is primarily on intra-domain routing while overlay routing is most commonly used to mask inter-domain routing inefficiencies.

\textbf{Interaction between traffic engineering and request redirection:} Recent work has studied the interaction between ISP traffic engineering and request redirection by content providers with geo-distribtued datacenters as well as P2P applications. Both analytical results \cite{Jiang2009,JohariGameTheory} and system implementations \cite{CATE,P4P} have shown that there is value for joint optimization of request redirection and traffic engineering, and cooperative strategies can help traffic engineering metrics while maintaining or improving user-perceived performance. Our work distinguishes in that  we model the three-way interaction between placement, redirection and routing and show that placement flexibility is more powerful degree of freedom than redirection or routing to improve traffic engineering metrics and to reduce user-perceived latency (Chapter \ref{ch:ncdn}).

@@

A content-dominated traffic and the presence of content at multiple network locations changes the traditional ISP traffic enginering problem. In such a network, a traffic matrix is no longer an effective representation of demand. It can be changed via application-layer adaptation without any underlying change in demand as we demonstrate in this section. Demand is better expressed as a \emph{content matrix} in which each entry represents the demand for a content at a particular network location. A traffic matrix implicitly assumes a fixed source and a fixed destination for each traffic flow. But, a content matrix expresses the fact that the demand for a content at a particular destination can be served from any set of source locations. As we explain using examples below, the flexibility of serving the demand from any location raises questions on the usefulness of link-utilization based metrics that are commonly used in traffic engineering (Section \ref{sec:bg-poormlu}) as well on the importance of traffic engineering itself (Example \ref{sec:bg-ncdn-interaction}). These and other related questions motivate our work on traffic-engineering in content dominated networks.

@@

\subsection{ISP network with content location diversity}
\label{sec:bg-locdiv}
We consider an ISP network that has little control over placement and redirection, but applications can leverage \emph{location diversity}, or  the ability to download content from multiple locatioons. We motivate why traffic engineering schemes must be evaluated while accounting for their interaction with location diversity, and why we need to look beyond link-utilization based metrics in evaluating traffic engineering schemes. These questions are addressed in Chapter \ref{ch:beyondmlu}.

@@

%We explain why link-utilization based metrics may not reflect user-perceived performance. Further, they are a poor metric to measure network capacity, i.e., the factor of surges in traffic demand that a network can tolerate, in a network with location diversity.

Link-utilization based metrics may not be a good predictor of user-perceived performance that depends on other factors such as propagtion delay, access link capacity and backbone link capacity also. Moreover, queuing-theoretic models show that only a high link utilization severely affects performance-critical metrics such as queuing delay and packet loss. For low to moderate link utilization that is common in today's ISPs, a reduction in link utilization may not yield a commensurate benefit in application peformance.

@@

Link-utilization based metrics such as MLU are poor metric for network capacity in the presence of location diversity. Unfortunately, as the example in Figure \ref{fig:3node-bg} shows, MLU is not a meaningful metric of capacity when application adaptation to location diversity determines the traffic matrix. 
Thus, location diversity necessitates a new capacity metric.

@@

\subsection{Network CDN}
\label{sec:bg-ncdn}
Network CDNs are a recent, potentialy transformative trend in which ISPs deploy CDNs on their infrastructures to deliver content to users on their network.  Unlike traditional ISPs and traditional CDNs, NCDNs enjoy full control over placement, redirection and routing on their networks. The interaction between placement and routing in an NCDN illustrated below motivates us to explore the benefit of joint optimization of placement, redirection and routing as well as to evaluate the relative importance of placement and routing in an NCDN. These questions are addressed in Chapter \ref{ch:ncdn}.

@@

To appreciate how placement can shape traffic in an NCDN, consider the simple example in Figure~\ref{fig:NetworkExample}. Node $C$ has an object in its cache that is requested by end-users at nodes $A$ and $D$. Suppose that one unit of traffic needs to be routed from $C$ to $A$ and $0.5$ units  from $C$ to $D$ to satisfy the demand for that object. The routing that achieves the minimum MLU of $0.5$ to serve the demanded object is shown in the figure. Note that the routing that achieves the MLU of 0.5 is not possible with a simple, \unplanned\ scheme like hop-count as that would route all the traffic demand from $C$ to $A$ via $B$, resulting in an MLU of 1. Thus, a (\planned) traffic engineering scheme is necessary to achieve an MLU of 0.5.

@@

On the other hand, NCDNs can shape the traffic demand matrix by using a judicious placement and redirection scheme. Suppose that there is some space left in the content server's cache at node $B$ to accommodate an additional copy of the demanded object. By creating an additional copy of the object at $B$, the traffic demand of $A$ can be satisfied from $B$ and the demand of $D$ from $C$ achieving an MLU of $0.125$. In this case, judicious content placement decreased the MLU by a factor of $4$. Even more interestingly, this best MLU can be achieved using a simple routing scheme like hop-count routing while also improving user-perceived latency (assuming that the latency of link $BA$ is lower than that of the two-hop paths from $C$ to $A$). While this is a toy example, it shows that content placement flexibility reduces network cost and enables simpler routing. This interaction between placement and routing motivates us address the following questions for real content workloads and network topologies.
